---
title: "ESM 244 Lab 2 Part 1"
subtitle: "Wrangling continued, PCA"
author: "Allison Horst"
date: "1/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)
```

## A. Get the data

Data: 

- California pollution burden (California Office of Environmental Health Hazard Assessment (OEHHA)’s CalEnviroScreen database, https://oehha.ca.gov/calenviroscreen/maps-data/download-data). See metadata (included as separate file). 
- California county demographics (from: CA census 2010)

Read it in: 
```{r}
ca_pb <- read_csv("ca_pollution_burden.csv")
ca_dem <- read_csv("ca_census_demographics_2010.csv")
```

## B. Do some cleaning 

1. For the pollution burden data: 

- Clean up the column headers
- Exclude any column that is a calculated percentile (contains 'percentile', 'perc', or 'pctl')

```{r}
ca_pb_nopct <- ca_pb %>% 
  clean_names() %>% 
  select(-contains("pctl")) %>% 
  select(-contains("perc")) %>% 
  select(-latitude, - longitude)
```

2. For the demographic data:

- Clean up column names

```{r}
ca_dem_clean <- ca_dem %>% 
  clean_names()
```


3. PCA for pollution burden indicator variables

First, starting with ca_pb_nopct:

**Note**: The pollution burden and population characteristic variables are aggregates (averages) of existing variables in the data frame, so we won't include those. That means we'll include columns:

- From `ozone:solid_waste`, and
- Frome `asthma:housing_burden`

First, just selecting those:
```{r}
ca_pb_subset <- ca_pb_nopct %>% 
  select(ozone:solid_waste, asthma:housing_burden)
```

We'll use the `prcomp` function:
- ?prcomp

```{r, eval = FALSE}

pb_pca <- prcomp(ca_pb_subset, scale = TRUE) # hmmm an error 

```

Look at the NA situation:
```{r}
summary(ca_pb_subset) # Max NAs in a variable: 242 (/8035)
```

A little aside: the `naniar` package for exploring missingness!
See: https://naniar.njtierney.com/

Use `naniar::gg_miss_var()` to plot the number of missings by variable:
```{r}
# Plot number of missings by variable
gg_miss_var(ca_pb_subset)
```

We can also explore NAs visually in other ways using the `VIM` package.

The `VIM::matrixplot()` function creates a heatmap, where values are shown on a continuous grayscale, and missings are in red:
```{r}
matrixplot(ca_pb_subset)

# We can also sort by a variable of interest. Let's say we want to sort by poverty:
matrixplot(ca_pb_subset, sortby = "poverty")
```

Let's say our conclusion is that there are missings, but not many (compared to the actual scope of the data). We'll only keep our complete cases (census tracts without any missings).

Use `tidyr::drop_na()` with no variables specified to keep complete cases across all variables:

```{r}
ca_pb_nona <- ca_pb_subset %>% 
  drop_na()

# Now check for NAs:
summary(ca_pb_nona)

# Or use `skimr::skim()`!
skim(ca_pb_nona)
```

Cool. No NAs, checked out missingness, NOW let's try PCA again: 

```{r}
my_ca_pca <- prcomp(ca_pb_nona, scale = TRUE)

# A nightmare! But what are we looking at? 
plot(my_ca_pca)
biplot(my_ca_pca)

# Hmmm let's try something else (this requires ggfortify): 
my_biplot <- autoplot(my_ca_pca, 
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE) +
  theme_minimal()
  

my_biplot


```

4. Join data by census tract (inner join)

```{r}

ca_df <- ca_dem_clean %>% 
  inner_join(ca_pb_nopct, by = c("census_tract_number" = "census_tract"))

```

Check it out, then get complete cases:
```{r}
ca_df_nona <- ca_df %>% 
  drop_na()
```

5. Make a new subset for PCA, that includes % white and elderly, and some interesting pollution burden & health indicators:

Like (you can choose a different set): 

- white_percent
- elderly_65_percent
- pm2_5
- pesticides
- traffic
- asthma
- cardiovascular_disease
- poverty

Make our subset:
```{r}
my_sub <- ca_df_nona %>% 
  select(white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)
```

Then run PCA: 
```{r}
my_dem_pca <- prcomp(my_sub, scale = TRUE)
biplot(my_dem_pca) # Noooo. We need better
```

Check it out a bit: 
```{r}
# Proportion of variance (& cumulative variance) explained by each PC
summary(my_dem_pca)

# Rotations (linear combinations for each PC):
my_dem_pca
```

Make a sweet biplot:
```{r}
my_dem_biplot <- autoplot(my_dem_pca, 
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE) +
  theme_minimal() +
  scale_y_continuous(limits = c(-0.05, 0.05))

my_dem_biplot
  
```

What are a few main things we can take out of this? What are the main correlations you notice? Are they in line with what you would expect, or is anything surprising? 

### End PCA section

